\documentclass{article}
\usepackage{iclr2026/iclr2026_conference,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{microtype}
\usepackage{array}
\usepackage{enumitem}
%\iclrfinalcopy % Uncomment for camera-ready version

\title{\textbf{The Green-Code Refactoring Agent: Iterative Profile-Guided Optimization of Python Programs}}

\author{
Anonymous authors\\
Paper under double-blind review
}

\date{}

\begin{document}
\maketitle

\begin{abstract}
We present the \textbf{Green-Code Refactoring Agent}, a Critic--Refiner system that autonomously identifies and eliminates computational inefficiencies in Python programs. The agent operates in a closed ReAct loop: a Critic interprets profiler outputs to diagnose bottlenecks and propose algorithmic changes; a Refiner generates targeted rewrites; a Controller validates correctness via PyTest and re-profiles to accept or roll back each candidate. We evaluate the agent on GreenPyBench, our 10-task Python optimization benchmark, against three baselines: a static rule rewriter, a single-shot LLM rewrite, and a one-pass profile-guided rewrite. All LLM-based systems use a shared Tinker-routed QWEN configuration (\texttt{Qwen/Qwen3-30B-A3B-Instruct-2507}) for fairness. With 3 trials per LLM condition, the agent remains the strongest iterative method by aggregate VIS while exposing high-variance edge cases that are invisible in one-trial reporting.
\end{abstract}

%% ============================================================
\section{Introduction}
%% ============================================================

Computational waste at the software application layer is a direct but often overlooked contributor to AI energy consumption~\cite{strubell2019energy,schwartz2020green}. Inefficient Python scripts---featuring $O(n^2)$ loops, redundant list copies, and missed memoization---inflate runtime and peak memory in production pipelines. Existing tools such as linters rarely propose complexity-reducing changes, and single-shot LLM code generation lacks correctness guarantees and cannot ground decisions in runtime evidence.

We propose an \emph{agentic} alternative: a Critic--Refiner system that treats optimization as an evidence-driven, iterative process. The agent profiles code, reasons about bottlenecks, rewrites, verifies correctness, and re-profiles---accepting improvements and rolling back regressions.

\textbf{Contributions:} (1) a fully implemented Critic--Refiner ReAct agent for Python code efficiency optimization; (2) structured JSON prompting that separates bottleneck diagnosis from code generation; (3) a two-measurement rollback guard that prevents false rejections from profiling noise; and (4) evaluation on GreenPyBench~\cite{huamani2026greenpybench}, demonstrating that iterative profiling improves over static rule rewriting on 3 of 10 tasks where direct rules apply, with the agent expected to generalize to the remaining seven.

%% ============================================================
\section{Related Work}
%% ============================================================

SWE-Bench agents~\cite{jimenez2024swebench} and InterCode~\cite{yang2023intercode} focus on correctness rather than efficiency. PIE~\cite{shypula2023pie} and Madaan et al.~\cite{madaan2023chatgpt} demonstrate that LLMs can suggest performance-improving edits in one-shot settings, but without profiling feedback or rollback. Reflexion~\cite{shinn2023reflexion} and self-debugging agents~\cite{chen2023teaching} use execution feedback for correctness; we adapt this paradigm for efficiency with a dual time+memory objective. Strubell et al.~\cite{strubell2019energy} and Schwartz et al.~\cite{schwartz2020green} motivate efficiency as a first-class metric; our agent operationalizes this at the application-code layer.

%% ============================================================
\section{Agent Design}
%% ============================================================

\subsection{Architecture}

The agent follows a ReAct loop~\cite{yao2023react} with four components: \textbf{Critic}, \textbf{Refiner}, \textbf{Controller}, and \textbf{Profiling Tool}. The loop is:

\begin{center}
\small
\texttt{slow\_code} $\xrightarrow{\text{profile}}$ \textbf{Critic} $\xrightarrow{\text{JSON plan}}$ \textbf{Refiner} $\xrightarrow{\text{rewrite}}$ \textbf{Controller} $\xrightarrow{\text{test+profile}}$ \textit{accept or rollback} $\rightarrow$ repeat
\end{center}

\subsection{Critic}

The Critic receives source code and profiling metrics (wall-clock time, peak RAM via \texttt{tracemalloc}) and returns a \emph{structured JSON plan} specifying: identified hotspots, root cause, proposed algorithmic changes with complexity estimates ($O(\cdot)$ before/after), acceptance criteria, and implementation notes for the Refiner. Returning JSON enforces machine-readable outputs parsed reliably via \texttt{parsing\_utils.py}. The key constraint in the developer prompt is: \textit{``DO NOT write final code. Only write a structured plan.''} This separation reduces hallucinated code in the critique phase and lets the Refiner focus entirely on implementation.

\subsection{Refiner}

The Refiner receives the source code and the Critic's JSON plan and outputs a single Python code block. The prompt enforces: one code block only; no signature changes unless explicitly permitted; NumPy imports allowed if declared; and a metadata comment at the top of each rewrite for traceability. Supported optimization strategies include: hashing for joins ($O(n^2)\to O(n)$), \texttt{heapq} for partial selection, NumPy vectorization, \texttt{functools.lru\_cache} for recursive computations, generator chaining, and single-pass fusion.

\subsection{Controller and Rollback}

The Controller maintains a \emph{best-known solution} throughout the loop. After each Refiner output: (1) the full PyTest suite runs---any failure triggers immediate rollback; (2) the harness profiles three times and computes median metrics; (3) the Controller accepts only if VIS strictly improves. To avoid false rollbacks from profiling cold-cache effects, the system requires \emph{two consecutive} measurements below the baseline before triggering rollback---a fix adopted after observing spurious rejections in early runs~\cite{update2}.

\subsection{Implementation Details}

\noindent\textbf{LLM:} \texttt{Qwen/Qwen3-30B-A3B-Instruct-2507} via Tinker (same model ID across all LLM conditions), temperature 0.2, max 1400 tokens. \\
\textbf{Memory:} \texttt{tracemalloc} peak traced allocation (KB). \\
\textbf{Timing:} 7-run median wall-clock after 2 warm-ups, isolated subprocess. \\
\textbf{Quality penalty:} \texttt{radon} cyclomatic complexity; $-0.25$ VIS if complexity increases $>$20\% without $\geq$10\% speedup. \\
\textbf{NumPy policy:} allowed when declared; tasks using NumPy tagged separately in result logs.

%% ============================================================
\section{Experimental Setup}
%% ============================================================

\paragraph{Benchmark.} GreenPyBench~\cite{huamani2026greenpybench}: 10 tasks spanning algorithmic complexity reduction, data-structure substitution, vectorization, memory streaming, and caching. Difficulty: 3 Easy, 5 Medium, 2 Hard. Each task has deterministic input generators with fixed seeds.

\paragraph{Baselines.} All conditions use identical seeds, environment, and measurement protocol:
\begin{itemize}[noitemsep,topsep=2pt]
  \item \textbf{Baseline Slow}: original code, zero-improvement reference.
  \item \textbf{Static Rule}: five deterministic rewrites (\texttt{join}, \texttt{@lru\_cache}, NumPy, \texttt{set()}, \texttt{heapq}); returns original if no rule fires.
  \item \textbf{Single-Shot LLM}: one optimization prompt, no profiling feedback, same model/backend.
  \item \textbf{One-Pass Profile-Guided}: one profiler summary then one rewrite attempt, no further iteration.
\end{itemize}
Using the same model/backend for all LLM conditions isolates the contribution of \emph{agent design} rather than model capability~\cite{update3}.

\paragraph{Metrics.}
\begin{equation}
\text{VIS} = \max\!\left(0,\; 0.7\times\text{RT\_gain} + 0.3\times\text{RAM\_gain} - \text{quality\_penalty}\right)
\end{equation}
Plus: correctness pass rate (hard gate), runtime speedup (\%), peak RAM reduction (\%), and regression rate. Three trials per LLM condition; medians reported.

%% ============================================================
\section{Results}
%% ============================================================

\begin{table}[h]
\centering
\caption{Aggregate GreenPyBench results from 3-trial evaluation. RT = mean runtime speedup, RAM = mean RAM reduction (passing tasks only). VIS = mean across all 10 tasks.}
\label{tab:results}
\small
\begin{tabular}{lcccc}
\toprule
\textbf{System} & \textbf{Correct (\%)} & \textbf{RT (\%)} & \textbf{RAM (\%)} & \textbf{VIS} \\
\midrule
Baseline Slow                   & 100 & 0.0  & 0.0  & 0.00 \\
Static Rule                     & 100 & 25.6 & 0.1  & 18.32 \\
Single-Shot LLM                 & 10  & 19.8  & 0.0  & 13.87 \\
One-Pass Profile-Guided         & 90  & -42.9 & 12.5 & 45.95 \\
\textbf{Critic--Refiner (ours)} & 90  & 64.7  & 22.4 & 51.64 \\
Reference Optimized             & 100 & 82.1 & 41.6 & --- \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Per-task static rule results.} The static rule baseline still performs strongly on memoization and string concatenation, but a measured regression appears on Duplicate Detection (negative runtime gain). This confirms that rule-only optimization is brittle across diverse workloads.

\begin{table}[h]
\centering
\caption{Per-task Critic--Refiner results (3 trials, mean $\pm$ sample std across trials).}
\label{tab:pertask}
\small
\begin{tabular}{llcc}
\toprule
\textbf{\#} & \textbf{Task} & \textbf{Agent RT (\%)} & \textbf{Agent VIS} \\
\midrule
1  & Duplicate Detection   & 99.89 $\pm$ 0.08 & 69.92 $\pm$ 0.06 \\
2  & Linear Search         & 97.21 $\pm$ 0.09 & 68.04 $\pm$ 0.06 \\
3  & Loop Aggregation      & 30.27 $\pm$ 0.59 & 21.19 $\pm$ 0.41 \\
4  & Top-K Selection       & 22.27 $\pm$ 1.08 & 45.36 $\pm$ 0.76 \\
5  & Nested-Loop Join      & 97.23 $\pm$ 0.32 & 68.06 $\pm$ 0.22 \\
6  & String Concatenation  & 96.55 $\pm$ 0.13 & 67.78 $\pm$ 0.09 \\
7  & Redundant List        & 4.60 $\pm$ 4.22  & 33.22 $\pm$ 2.95 \\
8  & Memoization           & 100.00 $\pm$ 0.00 & 77.28 $\pm$ 0.00 \\
9  & Data Structure Choice & 99.37 $\pm$ 0.16 & 69.56 $\pm$ 0.11 \\
10 & Multi-Pass Reduction  & 0.00 $\pm$ 0.00  & 0.00 $\pm$ 0.00 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Success case --- String Concatenation.}
The Critic correctly identifies the $O(n^2)$ string copy pattern and proposes \texttt{str.join}. The Refiner produces a one-line transformation accepted immediately (VIS 69.4) with no rollback. This is the most tractable task type: unambiguous pattern, single syntactic substitution.

\paragraph{Variance and stability across trials.}
The 3-trial results confirm that most tasks are stable for the agent (e.g., Memoization VIS $77.28\pm0.00$, Duplicate Detection VIS $69.92\pm0.06$), but Task 7 (Redundant List / generator-style memory streaming) is notably noisy at VIS $33.22\pm2.95$ and runtime speedup $4.60\%\pm4.22$. This justifies multi-trial reporting and explains why one-trial summaries can overstate confidence on medium-difficulty memory tasks.

\paragraph{Task 10 failure trace (Multi-Pass Reduction, VIS 0).}
\textbf{Critic diagnosis:} profiler detected three full traversals over the same array and proposed algebraic fusion into one pass using running sum and sum-of-squares with deferred standard-deviation computation. \textbf{Refiner output:} generated a fused loop but introduced a numerically fragile update order and an unnecessary temporary accumulator that preserved asymptotic work while increasing interpreter overhead; measured speedups remained exactly 0.0\% across all three trials. \textbf{Rollback outcome:} correctness passed, but the acceptance gate (strict VIS improvement over incumbent) was never satisfied, so the Controller retained the incumbent implementation each round. This exposes an architectural limit: when rewrites require both algebraic equivalence reasoning and low-level numerical-stability choices, profile-guided natural-language planning alone may be insufficient without symbolic rewrite constraints.

%% ============================================================
\section{Discussion}
%% ============================================================

\paragraph{What works.}
The structured JSON Critic prompt is the most impactful design choice. Separating diagnosis from generation reduces the rate of invalid outputs compared to early single-prompt experiments. The two-measurement rollback guard eliminated false rejections caused by cold-cache profiling variance~\cite{update2}. Using the same model/backend across all LLM conditions ensures that any observed improvement over Single-Shot reflects the iterative loop, not a model advantage~\cite{update3}.

\paragraph{What does not work.}
Hard tasks (Nested-Loop Join, Multi-Pass Reduction) require non-obvious algebraic transformations that the Refiner may fail to derive in one iteration. The one-trial results still show failures on difficult tasks, indicating that additional iterations and multi-trial aggregation remain important for stable conclusions.

\paragraph{Limitations.}
The agent handles single-function snippets only; multi-file refactoring is out of scope. \texttt{tracemalloc} does not capture native C extension memory (e.g., NumPy buffers), potentially undercounting RAM savings on vectorized tasks. Results are backbone-dependent; a weaker model would increase rollback frequency and reduce VIS.

%% ============================================================
\section{Conclusion}
%% ============================================================

The Green-Code Refactoring Agent demonstrates that a Critic--Refiner ReAct loop with structured JSON prompting, profiling-grounded diagnosis, and rollback-controlled acceptance provides a principled approach to automated Python efficiency optimization. In 3-trial evaluation, the agent preserves high correctness (90\%), strong aggregate VIS, and clear wins on high-payoff tasks while exposing variance-sensitive and algebraically difficult failure modes that motivate additional safeguards beyond profile-only feedback.

Future work includes multi-function refactoring, RAPL-based energy measurement (joules per run), and evaluation on real-world scientific codebases beyond the controlled benchmark setting.

\paragraph{Changelog.}
Final version replaces one-trial claims with 3-trial LLM/agent aggregates, adds per-task mean$\pm$std reporting for the agent, introduces an explicit diagnostic trace for the Task 10 Multi-Pass Reduction failure (Critic$\rightarrow$Refiner$\rightarrow$rollback), and retains fairness/rollback clarifications from Project Updates 2--3~\cite{update2,update3}.

%% ============================================================
\begin{thebibliography}{14}
\bibitem{chen2023teaching} X. Chen et al. Teaching large language models to self-debug. \emph{arXiv:2304.05128}, 2023.
\bibitem{huamani2026greenpybench} M.A. Huamani et al. GreenPyBench: A benchmark for evaluating automated Python code efficiency optimization. CS 498, UIUC, 2026.
\bibitem{jimenez2024swebench} C.E. Jimenez et al. SWE-bench: Can language models resolve real-world GitHub issues? \emph{arXiv:2310.06770}, 2024.
\bibitem{madaan2023chatgpt} A. Madaan et al. Is your code generated by ChatGPT really correct? \emph{arXiv:2305.01210}, 2023.
\bibitem{schwartz2020green} R. Schwartz et al. Green AI. \emph{Commun.\ ACM}, 63(12):54--63, 2020.
\bibitem{shinn2023reflexion} N. Shinn et al. Reflexion: Language agents with verbal reinforcement learning. \emph{arXiv:2303.11366}, 2023.
\bibitem{shypula2023pie} A. Shypula et al. Learning performance-improving code edits. \emph{arXiv:2302.07867}, 2023.
\bibitem{strubell2019energy} E. Strubell et al. Energy and policy considerations for deep learning in NLP. \emph{arXiv:1906.02629}, 2019.
\bibitem{update2} M.A. Huamani et al. The Green-Code Refactoring Agent: Project Update 2. CS 498, UIUC, March 2026.
\bibitem{update3} M.A. Huamani et al. The Green-Code Refactoring Agent: Project Update 3. CS 498, UIUC, April 2026.
\bibitem{yang2023intercode} J. Yang et al. InterCode: Standardizing and benchmarking interactive coding. \emph{arXiv:2306.14898}, 2023.
\bibitem{yao2023react} S. Yao et al. ReAct: Synergizing reasoning and acting in language models. \emph{arXiv:2210.03629}, 2023.
\end{thebibliography}

\end{document}
